Multiple imputation of incomplete multilevel data using Heckman selection models

### 2022-12-09 1.Paper/Paper.Rmd
Update toy example

corres: "* Johanna Munoz. \\email{j.munozavila@umcutrecht.nl}"
presentaddress: Julius Center for Health Sciences and Primary Care, University Medical Center Utrecht, Universiteitsweg 100, 3584 CG Utrecht, the Netherlands
authormark: Uthor \emph{et al}.
articletype: Research article
received: 2022-11-01
revised: 2022-12-01
accepted: 2023-01-01
abstract: "Missing data is a common problem in medical research, and is commonly addressed using multiple imputation. Although traditional imputation methods allow for valid statistical inference when data are missing at random (MAR), their implementation is problematic when the presence of missingness depends on unobserved information, i.e. the data are missing not at random (MNAR). Unfortunately, this MNAR situation is rather common, in observational studies, registries and other sources of real-world data. While several imputation methods have been proposed for addressing individual studies when data are MNAR, their application and validity in large datasets with clustering remains unclear. We therefore explored the consequence of MNAR data in hierarchical data in-depth, and proposed a novel multilevel imputation method for common missing patterns in clustered datasets. This method is based on the principles of Heckman selection models and adopts a two-stage meta-analysis approach to impute binary and continuous variables that may be outcomes or predictors and that are systematically or sporadically missing. After evaluating the proposed imputation model in simulated scenarios, we illustrate it use in a cross-sectional community survey to estimate the prevalence of malaria parasitemia in children aged 2-10 years in five subregions in Uganda."

keywords: Heckman model; IPDMA; Missing not at random; Selection models;Multiple imputation;
bibliography: bibfile.bib
output: 
  rticles::sim_article:
  fig_caption: yes
  longtable: yes
  doublespace: yes
longtable: true
header-includes:
   - \usepackage{mathtools}
   - \usepackage{float}
   - \usepackage{multicol}
   - \usepackage{graphicx}
     \newcommand{\indep}{\rotatebox[origin=c]{90}{$\models$}}
   - \usepackage{amsmath}
     \DeclareMathOperator\arctanh{arctanh}
   - \usepackage{booktabs}
   - \usepackage{longtable}
   - \usepackage{array}
   - \usepackage{multirow}
   - \usepackage{wrapfig}
   - \usepackage{float}
   - \usepackage{enumitem}
   - \usepackage{colortbl}
   - \usepackage{pdflscape}
   - \usepackage{tabu}
   - \usepackage{threeparttable}
   - \usepackage{makecell}     
   - \usepackage{xcolor}
   - \usepackage{array}
     \newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}p{#1}}

@@

# Introduction
Over the past few years, data sharing efforts have substantially increased, and researchers increasingly often have access to IPD from large combined datasets derived from electronic health records (EHR) or from multiple randomized or observable trials (i.e. in IPD meta-analysis, IPD-MA). For example, the clinical practice research datalink (CRPD) [@herrett2015] is an electronic health record dataset in the UK, which has been used in a variety of medical research, such as the evaluation of health policy and drug efficacy. A recent example of an IPD-MA is the emerging risk factor collaboration, [@theemergingriskfactorscollaboration2007] where data were combined from approximately 1.1 million individuals across 104 observational studies to investigate associations of cardiovascular diseases with several predictors. Individuals in these large datasets tend to be clustered in centres, countries or studies, where they have been subject to similar healthcare processes. Moreover, clusters may also differ in participant eligibility criteria, follow-up length, predictor and outcome definitions, or in the quality of applied measurement methods. Hence, differences (i.e.,heterogeneity) between clusters with respect to baseline covariates and outcomes, while the structure of the correlations between these variables is likely to be different across different clusters.

A usual problem is that such clustered datasets may contain  many incomplete variables. For example, in registry data it is common that test results are not available for all patients, as the decision to test may be at the discretion of the primary care physician or because the patient refuses to undergo testing. It is also possible that variables are systematically missing across clusters. For instance, in an IPD-MA, studies may have collected information on different variables. Missing values may thus appear for all participants of a study in the combined dataset. The presence of missing data can lead to loss of statistical power, imbalance in cluster size, bias in parameter estimates and therefore to erroneous conclusions as the analysis could be based on an unrepresentative sample.

To address the presence of missing data, it is important to consider  the proportion of missingness and the missing mechanism for each incomplete variable. Rubin (1976) [@rubin76] identified three missing mechanisms where the probability of missingness: 1) is constant (missing completely  at random; MCAR), 2) depends on observed data only (missing at random; MAR), or 3) depends on unobserved information even after conditioning on all observable variables (missing not at random; MNAR).  Traditional imputation methods are designed to address incomplete data sets where variables are MCAR or MAR. Their implementation is justified when there is not systematic difference between units with missing and  with complete data or when the missingness of a variable is strongly related to variables measured in the study.

Although it is possible to formally rule out MCAR, [@little1988]  it is impossible to assess whether data are MAR or MNAR as the observable data are not enough to test the assumptions of both mechanisms.[@enders2022]  For this reason, researchers often conveniently assume that data are MAR, or present results of a complete case analysis. In practice, however, unless missingness is artificially introduced by study design, e.g. when a test is only taken on patients with certain characteristics, missingness will often (at least partially) depend on unobserved information.

The missingness mechanism of a variable may be a mixture of MAR and MNAR, depending on the availability of data that may predict the missingness patterns. Therefore, the MAR mechanism may become more influential than MNAR when more information is recorded. For example, health professionals can decide which vital signs to measure on admission. When it is observed that low blood pressure (BP) measurements are not recorded, it could be assumed that this variable is MNAR, as the missigness of BP depends on its own value; however, BP could be considered as MAR variable if additional information is available to explain its missigness, such as the severity of the patient on admission.

@@

The MNAR mechanism is considered as non-ignorable, in this case it is necessary to specify information about the missingness process in addition to assumptions about the observed data. A popular approach to address MNAR is selection models such as the one proposed by Heckman (1976). [@heckman1976] Briefly, the Heckman selection model corrects for selection bias by estimating two linked equations: a outcome equation, where the missing variable is associated with predictors, and a selection equation, that accounts for the inclusion of observations in the sample.   An important feature of the Heckman selection model is that it does not assume data to be MNAR, so that it can also be used when data are MCAR or MAR. It therefore offers an appealing solution to incomplete data sets when the missingness mechanism is not precisely known.

@@

In section \ref{sec:heckman} we provide an introduction to the Heckman model and its estimation, and we extend it to a hierarchical setting. In section \ref{sec:methods} we define the main steps of the proposed imputation method. In section \ref{sec:simulation} we provide the settings and results of a simulation study to evaluate the performance of the proposed imputation method. In section \ref{sec:illustration} we illustrate the method using the survey information collected in different sub-districts in Uganda to estimate the prevalence of malaria in children. 
Finally in section \ref{sec:discussion} we summarize the results, outline limitations and propose future extension of our method.

@@

Generally the same variables can be used on the matrix of predictor variables $\boldsymbol{X^O_{i}}$ and $\boldsymbol{X^S_{i}}$. However, to avoid multicollinearity problems, [@puhani1997] it is recommended to include in $\boldsymbol{X^S_{i}}$ at least one variable that is not included in the outcome model. [@angrist2001] This variable is commonly known as an exclusion restriction variable (ERV), and should only be associated with the selection in the sample $r^*_{ij}$ (relevance condition) but not with the actual observation $y^*_{ij}$ (exclusion condition).[@gomes_etal20] The ERV meets by definition the relevance and exclusion conditions in order to provide independent information about the selection process and to facilitate the estimation of the Heckman model.

In the presence of selection bias,the aforementioned outcome equation will yield biased estimates of $\boldsymbol{\beta^O_{i}}$ if no efforts are made to adjust for the non-representativeness of the observed $\boldsymbol{X^O_{i}}$ and $\boldsymbol{y_{i}}$ values. For this reason, the Heckman model aims to jointly estimate the outcome and selection equation by defining a relation between their respective error distributions. For instance, Heckman’s original model [@heckman1976] assumes that residual terms have a bivariate normal distribution (BVN),

@@

Under the BVN assumption, the parameters of the Heckman model coefficients can be estimated using the two-step Heckman method (H2S) [@heckman1976] or the full information maximum likelihood method (FIML).[@amemiya1984] However, both methods may lead to inconsistent estimators when the true underlying distribution is not the BVN. [@gomes_etal20] To overcome this problem, other approaches have been proposed that relax the distribution assumptions, among them copula models. [@smith03] The copula approach uses a function, known as a copula, that joins the marginal distribution of the error terms of the selection and outcome equation which are specified separately. The Heckman model can be expressed with the following bivariate normal copula:
$$ F(\boldsymbol{r^*_{i}},\boldsymbol{y^*_{i}}) =\Phi\left(\boldsymbol{r^*_{i}}-\boldsymbol{X^S_{i}}\boldsymbol{\beta^S_{i}},\frac{\boldsymbol{y^*_{i}}-\boldsymbol{X^O_{i}}\boldsymbol{\beta^O_{i}}}{\sigma_i}, \rho_i\right),$$
where $\Phi$ is the bivariate standard normal cumulative distribution function.
Thus, to estimate the parameters of the Heckman method, it is sufficient to specify the marginal distributions of the error terms, and link them with a suitable copula function. In our imputation method we estimate the Heckman model using the copula method, as in real life data generating mechanism may deviate a BVN distribution. This facilitates the Heckman model estimation and makes it more robust to deviations of distributional assumptions.

@@

The Heckman model can be extended to hierarchical settings, i.e., in cases when individuals or sampling units are nested within clusters, as is the case in EHR or IPD. In this case, sample units from the same cluster are expected to share some characteristics (e.g., the distribution of variables and correlations between them, relationships between exposure and outcomes, missing processes).

Since most statistical analyses assume that the sample units are independent of each other, more complex hierarchical models dealing with nested data are required.  This hierarchical complexity must not only be taken into account in the analysis model, but also when dealing with missing data, thus requiring imputation models that are congenial to the analysis model, i.e. that make the same assumptions about the data.

@@

Briefly, in a first stage  we estimate $\boldsymbol{\theta_i}=\{\boldsymbol{\beta_i^O},\boldsymbol{\beta_i^S},\sigma_i,\rho_i\}$ the cluster specific parameters of the Heckman model. That is, we estimate separately for each of the $N$ clusters,the parameters of the two equations in each study, the outcome model and the selection model. In the second stage, we combine all $\boldsymbol{\theta_i}$  by using a random effects multivariate meta-analysis model.

@@

Initially, we use the copula method to estimate the set of cluster-specific parameters, $\boldsymbol{\widehat{\theta_i}}=\boldsymbol{\{\widehat{\beta^O_i}}, \boldsymbol{\widehat{\beta^S_i}}, \widehat{\sigma_i}, \widehat{\rho_i}\}$, using all $j$ units with observable measurements $y_{ij}^{obs}$ within each cluster $i$. 
The Heckman model is estimated with the \textbf{gjrm} function of the GJRM R package, [@radiceGJRMGeneralisedJoint2021] under the bivariate model with the nonrandom sample selection (BSS) specification, from which we obtain not only the parameters' point estimates $\boldsymbol{\widehat{\theta_i}}$, but also their corresponding $\boldsymbol{\widehat{S(\theta_i)}}$ within-cluster variance-covariance matrix.

@@

Our imputation model can then be used in the imputation of any incomplete variable in a dataset following an MNAR mechanism, even if it is an outcome, predictor or auxiliary variable of the main analysis model. Furthermore, as implemented according to the MICE framework, the imputation method could be used simultaneously with other imputation methods available in MICE or in add-in packages such as MICEMD [@audigier2021], to impute datasets with multiple incomplete variables that differ in type and missing mechanism.

@@

## Aim
We designed a simulation study aimed to compare the performance of alternative methods for imputing a single missing outcome variable in a hierarchical dataset, where the missingness follows a MNAR mechanism. In our scenarios we considered systematically missingness,i.e. the case some of the studies did not report hte variable at all.

@@

We generated the data from a Heckman selection model with bivariate normal distribution error terms. For simplicity we started from "basic scenario" where the database collected information from $N=10$ clusters of $n_i=1000$  individuals. However, we altered both $N$ and $n_i$ in sensitivity analyses (Table \ref{tab:scenarios}). 
```{r scenarios,  echo = FALSE, warning = FALSE,cache = TRUE}
data2 <- read.table(stringsAsFactors = FALSE, header = TRUE, sep="/", text =
'Scenario/	Incomplete variable/	Rho/	N; $n_i$/	Missing process
Base/	Continuous/	0.6/	10;1000/	Heckman(BVN)
M(N)AR/Continuous, Binary/	MAR:0, MNAR:0.3,0.6,0.9/	10;1000/Heckman(BVN)
Size and cluster number/	Continuous/	0.6/	10;50,   10;100,   10;1000,  50;1000,  100;1000/	Heckman (BVN)
Distribution deviations/	Continuous/	0.6/	10;1000	/Heckman(BVN), Heckman(t-skew), Explicit'
)
options(kableExtra.latex.load_packages = FALSE) 
library(kableExtra)
kbl(data2, booktabs = T, align="l",
    caption = "Data generation scenarios",
    col.names = c("Scenario","Incomplete variable", "$\\rho$", "$N;n_i$","Missing process"),
    row.names=FALSE,escape = FALSE)%>%
    column_spec(c(1), width = "2,5cm")%>%
    column_spec(c(2,4), width = "1,5cm")%>%
    column_spec(c(3), width = "3cm")%>%
    column_spec(c(5), width = "2.7cm")%>%
  kable_styling(latex_options = c("striped"))
```
For each dataset, we generated $X_{1i}$ a treatment indicator variable from a Bernoulli distribution with a probability of treatment on each cluster equal to 0.6. Next, we simulated the mean of two continuous covariates from a multivariate normal distribution $\mu_h\sim N(\begin{psmallmatrix}0\\0\end{psmallmatrix},\begin{psmallmatrix}0.2 & 0.015\\0.015 & 0.2\end{psmallmatrix})$, with $h=\{2,3\}$. We then simulated for each cluster a baseline covariate $X_{2i}\sim N(\mu_2,1)$ and a exclusion restriction variable  $X_{3i} \sim N(\mu_3,0.5)$.

@@

## Analysis Methods

@@

- \textbf{Complete case analysis (CCA):} We removed all patients with missing observations. 
- \textbf{1l.Heckman:} Multiple imputation based on the Heckman model without no study specification, following the imputation method proposed by Galimard et al.(2016). [@galimard_etal16]
- \textbf{2l.MAR}: Multiple imputation assuming MAR for hierarchical datasets, we used the multilevel imputation model ( 2l.2stage.norm and 2l.2stage.bin) from the micemd R package, [@audigier2021] which are described in Audigier et al. (2018) paper. [@audigier_etal18]

@@

After the imputation procedure, we estimated the following (generalized) mixed linear effect model using the *lmer()* function from the *lme4* R package.[@bates2022]
$y_{i}=\beta_i^OX_i^O+\epsilon^O_{i}$
In case of missing binary variable, we used the same matrix of predictors but on a binary model estimated with the *glmer()* function from the *lme4* R package. Then, we pooled the estimates of the $\beta_i^O$ and the variance of the random effect and residual errors of the multiple imputed datasets according to Rubin's rule, [@rubin1987] over which we calculated the performance measures on the estimands.

To calculate the coverage of the parameter coefficients' \95% confidence intervals (CI), we estimate CI with the Wald method. Although it is possible to obtain CI through the profile or bootstrap method for the variances of the random effects, we prefer not to estimate them (and hence the coverage) for the random effects parameters due to computational time.

@@

We calculated the following measures, usually employed to evaluate imputation methods, [@buuren18c] according to the formulas provided in Morris et al.(2019): [@morris2019]

@@

The Heckman 2L imputation method is available in the MICEMD R package [@audigier2021] (as \textbf{mice.2l.heckman()}) and also on the github repository \url{https://github.com/johamunoz/Statsmed_Heckman} where you can also find all the codes accompanying this paper and a toy example that explains how to implement the method in mice.

@@

### Descriptive statistics of the simulated datasets
We generated data sets of 10 clusters of 1000 patients each in each scenario. For example, for the base scenario (Table \ref{tab:scenarios}) in which the error terms followed a normal distribution, out of the 500 datasets generated, we obtained that on average 60.8\% of the Y response was missing, with the lowest missing percentage being 26.84\% and the maximum being 73.67\%.  At the cluster level, looking only at the sporadic missing clusters, we found an average of 35.25\% missing values, but there were clusters with no missing data at all and up to 98\% missing data in the outcome. Processing times differ in the imputation methods evaluated, the 1l. Heckman takes an average of 10 seconds, the 2l.MAR takes 1.5 and the 2l. Heckman takes about 18 seconds. It should be noted that the time of the last two approaches depends strongly on the number of clusters included and the sample size in the data set. (See Annex excel file)

@@

For this example, we used data accessed directly from CliniEpiDB, [@staedke] where data were collected from 5195 households with verified consent, inhabited by 11137 residents aged 2-10 years. Blood samples were only taken from 8846 children, as 69 were excluded from the study due to lack of consent and 2222 were not present at the time of the survey. Although the original data set consists of 164 variables, here we only consider the variables described in Table \ref{tab:descriptive}, which were used as predictors in the imputation model and are fully observed in the dataset. In this dataset, the parasitaemia test is an incomplete binary result (1=positive test, 0=negative test), which is missing 21\% across the whole dataset.

@@

Under the MCAR assumption, prevalence was calculated on the basis of the recorded tests, i.e., we only included patients with a test result. Under the MAR assumption, the test values of children who were not present during the survey were imputed with the 2l.2stage.bin method of the MICEMD package, where the community was taken as the cluster and the following factors previously associated with parasitemia were used as predictors in the imputation model: sex, bednet ( indicator of whether only two or fewer persons share a mosquito bed net).  In addition, we included age as a power 3 spline function, the cluster-level Log10 mean of the number of female anopheline mosquitoes per household estimated from the entomological survey, and the household wealth index from principal components analysis calculated specifically for the surveyed households.

@@

In our simulations we observed that the imputation method we proposed id optimal for the imputation of continuous and binary missing variables that follow a MNAR mechanism according to the Heckman model and that come from multilevel data, such data commonly used in the IPDMA.

>>>>

corres: "* Johanna Munoz. \\email{j.munozavila@umcutrecht.nl}"
presentaddress: Julius Center for Health Sciences and Primary Care, University Medical Center Utrecht, Universiteitsweg 100, 3584 CG Utrecht, the Netherlands
authormark: Uthor \emph{et al}.
articletype: Research article
received: 2022-11-01
revised: 2022-12-01
accepted: 2023-01-01
abstract: "Missing data is a common problem in medical research, and is commonly addressed using multiple imputation. Although traditional imputation methods allow for valid statistical inference when data are missing at random (MAR), their implementation is problematic when the presence of missingness depends on unobserved variables, i.e. the data are missing not at random (MNAR). Unfortunately, this MNAR situation is rather common, in observational studies, registries and other sources of real-world data. While several imputation methods have been proposed for addressing individual studies when data are MNAR, their application and validity in large datasets with multilevel structure remains unclear. We therefore explored the consequence of MNAR data in hierarchical data in-depth, and proposed a novel multilevel imputation method for common missing patterns in clustered datasets. This method is based on the principles of Heckman selection models and adopts a two-stage meta-analysis approach to impute binary and continuous variables that may be outcomes or predictors and that are systematically or sporadically missing. After evaluating the proposed imputation model in simulated scenarios, we illustrate it use in a cross-sectional community survey to estimate the prevalence of malaria parasitemia in children aged 2-10 years in five subregions in Uganda."

keywords: Heckman model; IPDMA; Missing not at random; Selection models; Multiple imputation;
bibliography: bibfile.bib
output: 
  rticles::sim_article:
  fig_caption: yes
  longtable: yes
  doublespace: yes
longtable: true
header-includes:
   - \usepackage{mathtools}
   - \usepackage{float}
   - \usepackage{multicol}
   - \usepackage{graphicx}
     \newcommand{\indep}{\rotatebox[origin=c]{90}{$\models$}}
   - \usepackage{amsmath}
     \DeclareMathOperator\arctanh{arctanh}
   - \usepackage{booktabs}
   - \usepackage{longtable}
   - \usepackage{array}
   - \usepackage{multirow}
   - \usepackage{wrapfig}
   - \usepackage{float}
   - \usepackage{enumitem}
   - \usepackage{colortbl}
   - \usepackage{pdflscape}
   - \usepackage{tabu}
   - \usepackage{threeparttable}
   - \usepackage{makecell}     
   - \usepackage{xcolor}
   - \usepackage{array}
     \newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}p{#1}}

@@

# Introduction
Over the past few years, data sharing efforts have substantially increased, and researchers increasingly often have access to IPD from large combined datasets derived from electronic health records (EHR) or from multiple randomized or observable trials (i.e. in IPD meta-analysis, IPD-MA). For example, the clinical practice research datalink (CRPD) [@herrett2015] is an electronic health record dataset in the UK, which has been used in a variety of medical research, such as the evaluation of health policy and drug efficacy. A recent example of an IPD-MA is the emerging risk factor collaboration[@theemergingriskfactorscollaboration2007], where data were combined from approximately 1.1 million individuals across 104 observational studies to investigate associations of cardiovascular diseases with several predictors. Individuals in these large datasets tend to be clustered in centres, countries or studies, where they have been subject to similar healthcare processes. Moreover, clusters may also differ in participant eligibility criteria, follow-up length, predictor and outcome definitions, or in the quality of applied measurement methods. Hence, heterogeneity between clusters with respect to baseline covariates and outcomes, while the structure of the correlations between these variables is likely to be different across different clusters.

A usual problem is that such clustered datasets may contain  many incomplete variables. For example, in registry data it is common that test results are not available for all patients, as the decision to test may be at the discretion of the primary care physician or because the patient refuses to undergo testing. It is also possible that variables are systematically missing across clusters.[@resche-rigon2013] For instance, in an IPD-MA, studies may have collected information on different variables. Missing values may thus appear for all participants of a study in the combined dataset. The presence of missing data can lead to loss of statistical power, imbalance in cluster size, bias in parameter estimates and therefore to erroneous conclusions as the analysis could be based on an unrepresentative sample.

To address the presence of missing data, it is important to consider the missing mechanism for each incomplete variable. Rubin (1976) [@rubin76] identified three missing mechanisms where the probability of missingness: 1) is independent from observed or missing values (missing completely  at random; MCAR), 2) depends on observed data only (missing at random; MAR), or 3) depends on unobserved information even after conditioning on all observable variables (missing not at random; MNAR).  Traditional imputation methods are designed to address incomplete data sets where variables are MCAR or MAR. Their implementation is justified when there is not systematic difference between units with missing and  with complete data or when the missingness of a variable is strongly related to variables measured in the study.

<!-- Although it is possible to formally rule out MCAR[@little1988], it is impossible to assess whether data are MAR or MNAR as the observable data are not enough to test the assumptions of both mechanisms.[@enders2022]  For this reason, researchers often conveniently assume that data are MAR, or present results of a complete case analysis. In practice, however, unless missingness is artificially introduced by study design, e.g. when a test is only taken on patients with certain characteristics, missingness will often (at least partially) depend on unobserved information.  -->

<!-- The missingness mechanism of a variable may be a mixture of MAR and MNAR, depending on the availability of data that may predict the missingness patterns. Therefore, the MAR mechanism may become more influential than MNAR when more information is recorded. For example, health professionals can decide which vital signs to measure on admission. When it is observed that low blood pressure (BP) measurements are not recorded, it could be assumed that this variable is MNAR, as the missigness of BP depends on its own value; however, BP could be considered as MAR variable if additional information is available to explain its missigness, such as the severity of the patient on admission. -->

@@

Modelling data under MNAR mechanism implies to specify information about the missingness process in addition to assumptions about the observed data. A popular approach to address MNAR mechanism is selection models such as the one proposed by Heckman (1976). [@heckman1976] Briefly, the Heckman selection model corrects for selection bias by estimating two linked equations: a outcome equation, where the missing variable is associated with predictors, and a selection equation, that accounts for the inclusion of observations in the sample.   An important feature of the Heckman selection model is that it does not assume data to be MNAR, so that it can also be used when data are MCAR or MAR. It therefore offers an appealing solution to incomplete data sets when the missingness mechanism is not precisely known.

@@

In section \ref{sec:heckman} we provide an introduction to the Heckman model and its estimation, and we extend it to a hierarchical setting. In section \ref{sec:methods} we define the main steps of the proposed imputation method. In section \ref{sec:simulation} we provide the settings and results of a simulation study to evaluate the performance of our imputation method. In section \ref{sec:illustration} we illustrate the method using the survey information collected in different sub-districts in Uganda to estimate the prevalence of malaria in children. 
Finally, in section \ref{sec:discussion} we summarize the results, outline limitations and propose future extension of our method.

@@

Generally the same variables can be used on the matrix of predictor variables $\boldsymbol{X^O_{i}}$ and $\boldsymbol{X^S_{i}}$. However, to avoid multicollinearity problems[@puhani1997], it is recommended to include in $\boldsymbol{X^S_{i}}$ at least one variable that is not included in the outcome model. [@angrist2001] This variable is commonly known as an exclusion restriction variable (ERV), and should only be associated with the selection in the sample $r^*_{ij}$ (relevance condition) but not with the actual observation $y^*_{ij}$ (exclusion condition).[@gomes_etal20] The ERV meets by definition the relevance and exclusion conditions in order to provide independent information about the selection process and to facilitate the estimation of the Heckman model.

In the presence of selection bias, the aforementioned outcome equation will yield biased estimates of $\boldsymbol{\beta^O_{i}}$ if no efforts are made to adjust for the non-representativeness of the observed $\boldsymbol{X^O_{i}}$ and $\boldsymbol{y_{i}}$ values. For this reason, the Heckman model aims to jointly estimate the outcome and selection equation by defining a relation between their respective error distributions. For instance, Heckman’s original model [@heckman1976] assumes that residual terms have a bivariate normal distribution (BVN),

@@

Under the BVN assumption, the parameters of the Heckman model coefficients can be estimated using the two-step Heckman method (H2S) [@heckman1976] or the full information maximum likelihood method (FIML).[@amemiya1984] However, both methods may lead to inconsistent estimators when the true underlying distribution is not the BVN. [@gomes_etal20] To overcome this problem, other approaches have been proposed that relax the distribution assumptions, among them copula models. [@smith03] The copula approach uses a function, known as a copula, that joins the marginal distribution of the error terms of the selection and outcome equation which are specified separately. The Heckman model can be expressed with the following bivariate normal copula:
$$ F(\boldsymbol{r^*_{i}},\boldsymbol{y^*_{i}}) =\Phi\left(\boldsymbol{r^*_{i}}-\boldsymbol{X^S_{i}}\boldsymbol{\beta^S_{i}},\frac{\boldsymbol{y^*_{i}}-\boldsymbol{X^O_{i}}\boldsymbol{\beta^O_{i}}}{\sigma_i}, \rho_i\right),$$
where $\Phi$ is the bivariate standard normal cumulative distribution function.
Thus, to estimate the parameters of the Heckman method, it is sufficient to specify the marginal distributions of the error terms, and link them with a suitable copula function. In our imputation method we estimate the Heckman model using the copula method, as in real life data generating mechanism may deviate a BVN distribution. <!-- This facilitates the Heckman model estimation and makes it more robust to deviations of distributional assumptions. -->

@@

 The Heckman model can be extended to hierarchical settings, i.e., in cases when individuals or sampling units are nested within clusters, as is the case in EHR or IPD. <!--In this case, sample units from the same cluster are expected to share some characteristics (e.g., the distribution of variables and correlations between them, relationships between exposure and outcomes, missing processes).

Since most statistical analyses assume that the sample units are independent of each other, more complex hierarchical models dealing with nested data are required. --> This hierarchical complexity must not only be taken into account in the analysis model, but also when dealing with missing data, thus requiring imputation models that are congenial to the analysis model, i.e. that make the same assumptions about the data.

@@

Briefly, in a first stage  we estimate $\boldsymbol{\theta_i}=\{\boldsymbol{\beta_i^O},\boldsymbol{\beta_i^S},\sigma_i,\rho_i\}$, for all $i$ between 1 and $N$ clusters, the cluster specific parameters of the Heckman model. That is, we estimate separately for each of the clusters, the parameters of the two equations in each study, the outcome model (\ref{eq:1}) and the selection model (\ref{eq:2}). In the second stage, we combine all $\boldsymbol{\theta_i}$  by using a random effects multivariate meta-analysis model.

@@

Initially, we use the copula method to estimate the set of cluster-specific parameters, $\boldsymbol{\widehat{\theta_i}}=\boldsymbol{\{\widehat{\beta^O_i}}, \boldsymbol{\widehat{\beta^S_i}}, \widehat{\sigma_i}, \widehat{\rho_i}\}$, using all $j$ units with observable measurements $y_{ij}^{obs}$ within each cluster $i$. 
The Heckman model is estimated with the \textbf{gjrm} function of the GJRM R package[@radiceGJRMGeneralisedJoint2021],  under the bivariate model with the nonrandom sample selection (BSS) specification, from which we obtain not only the parameters' point estimates $\boldsymbol{\widehat{\theta_i}}$, but also their corresponding $\boldsymbol{\widehat{S(\theta_i)}}$ within-cluster variance-covariance matrix.

@@

Our imputation model can then be used in the imputation of any incomplete variable in a dataset following an MNAR mechanism, even if it is an outcome, predictor or auxiliary variable of the main analysis model. Furthermore, as implemented according to the MICE framework, the imputation method could be used simultaneously with other imputation methods available in MICE or in add-in packages such as micemd [@audigier2021], to impute datasets with multiple incomplete variables that differ in type and missing mechanism.

@@

## Aim
We designed a simulation study aimed to compare the performance of alternative methods for imputing a single missing outcome variable in a hierarchical dataset, where the missingness follows a MNAR mechanism. In our scenarios we considered systematically missingness.

@@

We generated the data from a Heckman selection model with bivariate normal distribution error terms. For simplicity we started from "basic scenario" where the database collected information from $N=10$ clusters of $n_i=1000$  individuals. However, we altered both $N$ and $n_i$ in sensitivity analyses (Table \ref{tab:scenarios}). 
```{r scenarios,  echo = FALSE, warning = FALSE,cache = TRUE}
data2 <- read.table(stringsAsFactors = FALSE, header = TRUE, sep="/", text =
'Scenario/	Incomplete variable/	Rho/	N; $n_i$/	Missing process
Base/	Continuous/	0.6/	10;1000/	Heckman (BVN)
M(N)AR/Continuous, Binary/ 0 (MAR), 0.3,0.6,0.9 (MNAR)/	10;1000/Heckman (BVN)
Size and cluster number/	Continuous/	0.6/	10;50,   10;100,   10;1000,  50;1000,  100;1000/	Heckman (BVN)
Distribution deviations/	Continuous/	0.6/	10;1000	/Heckman (BVN), Heckman (t-skew), Explicit'
)
options(kableExtra.latex.load_packages = FALSE) 
library(kableExtra)
kbl(data2, booktabs = T, align="l",
    caption = "Data generation scenarios",
    col.names = c("Scenario","Incomplete variable", "$\\rho$", "$N;n_i$","Missing process"),
    row.names=FALSE,escape = FALSE)%>%
    column_spec(c(1), width = "2,5cm")%>%
    column_spec(c(2,4), width = "1,5cm")%>%
    column_spec(c(3), width = "2,9cm")%>%
    column_spec(c(5), width = "2.7cm")%>%
  kable_styling(latex_options = c("striped"))
```
For each dataset, we generated $X_{1i}$ a treatment indicator variable from a Bernoulli distribution with a probability of treatment on each cluster equal to 0.6. Next, we simulated the mean of two continuous covariates from a multivariate normal distribution $\mu_h\sim N(\begin{psmallmatrix}0\\0\end{psmallmatrix},\begin{psmallmatrix}0.2 & 0.015\\0.015 & 0.2\end{psmallmatrix})$, with $h=\{2,3\}$. We then simulated for each cluster a baseline covariate $X_{2i}\sim N(\mu_2,1)$ and a exclusion restriction variable  $X_{3i} \sim N(\mu_3,0.5)$.

@@

## Imputation methods

@@

- \textbf{Complete case analysis (CCA):} We removed all patients with missing observations. 
- \textbf{1l.Heckman:} Multiple imputation based on the Heckman model without no study specification, following the imputation method proposed by Galimard et al.(2016). [@galimard_etal16]
- \textbf{2l.MAR}: Multiple imputation assuming MAR for hierarchical datasets, we used the multilevel imputation model ( 2l.2stage.norm and 2l.2stage.bin) from the micemd R package[@audigier2021], which are described in Audigier et al. (2018) paper. [@audigier_etal18]

@@

After the imputation procedure, we estimated the following (generalized) mixed linear effect model using the *lmer()* function from the *lme4* R package.[@bates2022]
$y_{i}=\beta_i^OX_i^O+\epsilon^O_{i}$
In case of missing binary variable, we used the same matrix of predictors but on a binary model estimated with the *glmer()* function from the *lme4* R package. Then, we pooled the estimates of the $\beta_i^O$ and the variance of the random effect and residual errors of the multiple imputed datasets according to Rubin's rule[@rubin1987], over which we calculated the performance measures on the estimands.

To calculate the coverage of the parameter coefficients' 95\% confidence intervals (CI), we estimate CI with the Wald method. <!-- Although it is possible to obtain CI through the profile or bootstrap method for the variances of the random effects, we prefer not to estimate them (and hence the coverage) for the random effects parameters due to computational time.  -->

@@

We calculated the following measures, usually employed to evaluate imputation methods[@buuren18c], according to the formulas provided in Morris et al.(2019)[@morris2019]:

@@

The Heckman 2L imputation method is available in the micemd R package [@audigier2021] (as \textbf{mice.2l.heckman()}) and also on the github repository \url{https://github.com/johamunoz/Statsmed_Heckman} where you can also find all the codes accompanying this paper and a toy example that explains how to implement the method in mice.

@@

<!-- ### Descriptive statistics of the simulated datasets -->
<!-- We generated data sets of 10 clusters of 1000 patients each in each scenario. For example, for the base scenario (Table \ref{tab:scenarios}) in which the error terms followed a normal distribution, out of the 500 datasets generated, we obtained that on average 60.8\% of the Y response was missing, with the lowest missing percentage being 26.84\% and the maximum being 73.67\%.  At the cluster level, looking only at the sporadic missing clusters, we found an average of 35.25\% missing values, but there were clusters with no missing data at all and up to 98\% missing data in the outcome. Processing times differ in the imputation methods evaluated, the 1l. Heckman takes an average of 10 seconds, the 2l.MAR takes 1.5 and the 2l. Heckman takes about 18 seconds. It should be noted that the time of the last two approaches depends strongly on the number of clusters included and the sample size in the data set. (See Annex excel file) -->

@@

For this example, we used data accessed directly from ClinEpiDB[@staedke], where data were collected from 5195 households with verified consent, inhabited by 11137 residents aged 2-10 years. Blood samples were only taken from 8846 children, as 69 were excluded from the study due to lack of consent and 2222 were not present at the time of the survey. Although the original data set consists of 164 variables, here we only consider the variables described in Table \ref{tab:descriptive}, which were used as predictors in the imputation model and are fully observed in the dataset. In this dataset, the parasitaemia test is an incomplete binary result (1=positive test, 0=negative test), which is missing 21\% across the whole dataset.

@@

Under the MCAR assumption, prevalence was calculated on the basis of the recorded tests, i.e., we only included patients with a test result. Under the MAR assumption, the test values of children who were not present during the survey were imputed with the 2l.2stage.bin method of the micemd package, where the community was taken as the cluster and the following factors previously associated with parasitemia were used as predictors in the imputation model: sex, bednet ( indicator of whether only two or fewer persons share a mosquito bed net).  In addition, we included age as a power 3 spline function, the cluster-level Log10 mean of the number of female anopheline mosquitoes per household estimated from the entomological survey, and the household wealth index from principal components analysis calculated specifically for the surveyed households.

@@

In our simulations we observed that the imputation method we proposed is optimal for the imputation of continuous and binary missing variables that follow a MNAR mechanism according to the Heckman model and that come from multilevel data, such data commonly used in the IPDMA.
